A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-12 sections/Overview.tex
move algorithm paragraphs from 2.1 to 2.2
DK algorithm description

\paragraph{Metatheory} % Talk about the lemmas that are false, show counterexample.
We have tried to follow DK's proof for formalization,
but failed due to incorrect proof techniques or logic errors.
The proof of Lemma 19 (Extension Equality Preservation) applies a wrong
induction hypothesis for the ${\longrightarrow}\mathtt{Uvar}$ case.
Fortunately the lemma could be proven without the well-formedness conditions
$\Gm\vdash A$ and $\Gm\vdash B$.
Similar problem happens for Lemma 14 (Subsumption), where many applications of
induction hypotheses are not correct.
For manual proofs, induction hypotheses are not automatically generated by programs,
thus are easily misused.

@@

%-------------------------------------------------------------------------------
\subsection{DK's Algorithm}
\jimmy{TODO: some of the text above about DK's algorithm may be moved here}

>>>>

%-------------------------------------------------------------------------------
\subsection{DK's Algorithm}

The DK algorithm revolves around their notion of \emph{algorithmic context}.
\[
\begin{array}{l@{\qquad}lcl}
\text{Algorithmic Contexts}\qquad&\Gamma,\Delta,\Theta &::=&\quad \nil \mid
\Gamma, a \mid \Gamma, \hat{\alpha} \mid \Gamma, \hat{\alpha} = \tau \mid
\Gamma, \blacktriangleright_{\hat{\alpha}}
\end{array}
\]
In addition to the regular (universally quantified) type variables $a$, the
algorithmic context also contains \emph{existential} type variables
$\hat{\alpha}$. These are placeholders for monotypes $\tau$ that are still to
be determined by the inference algorithm. When the existential variable is
``solved'', its entry in the context is replaced by the assignment
$\hat{\alpha} = \tau$. With the notation $[\Gamma]$ an algorithmic context
$\Gamma$ denotes the substitution that replaces all its solved existential type
variables with their solution.

All algorithmic judgments thread an algorithmic context. They have the form
$\Gamma \vdash \ldots \dashv \Delta$, where $\Gamma$ is the input context and
$\Delta$ is the output context:
$\Gamma \vdash A \le B \dashv \Delta$  for subtyping, 
$\Gamma \vdash e \Leftarrow A \dashv \Delta$  for type checking, and so on. 
The output context is a functional update of the input context that records newly
introduced existentials and solutions.

Solutions are incrementally propagated by applying the algorithmic output
context of a previous task as substitution to the next task. This can be seen
in the subsumption rule:
\[
\inferrule*
  {\Gamma \vdash e \Rightarrow A \dashv \Theta \\ 
   \Theta \vdash [\Theta]A \le [\Theta]B \dashv \Delta
  }
  { \Gamma \vdash e \Leftarrow B \dashv \Delta}
\]
The inference task yields an output context $\Theta$ which is applied as a substitution
to the types $A$ and $B$ before performing the subtyping check to propagate any solutions
of existential variables that appear in $A$ and $B$.

The sequential order of entries in the algoritmic context, in combination with
the threading of contexts,  does not perfectly capture the scoping of all
existential variables. For this reason the DK algorithm uses scope markers
$\blacktriangleright_{\hat{\alpha}}$ in a few places. An example is given in the following
rule:
\[
\inferrule*
  {\Gamma,\blacktriangleright_{\hat{\alpha}}, \hat{\alpha} \vdash [\hat{\alpha}/a]A \le B \dashv \Delta,\blacktriangleright_{\hat{\alpha}},\Theta}
  {\Gamma \vdash \forall a. A \le B \dashv \Delta}
\]
To indicate that the scope of $\hat{\alpha}$ is local to the subtyping check
$[\hat{\alpha}/a]A \le B$, the marker is pushed onto its input stack and popped
from the output stack together with the subsequent part $\Theta$, which may
refer to $\hat{\alpha}$. (Remember that later entries may refer to earlier
ones, but not vice versa.) This way $\hat{\alpha}$ does not escape its scope.

At first sight, the DK algorithm would seem a good basis for mechanisation. After
all it comes with a careful description and extensive manual proofs.
Unfortunately, we ran into several obstacles that have prompted us to formulate
a different, more mechanisation-friendly algorithm.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Broken Metatheory} % Talk about the lemmas that are false, show counterexample.
While going through the manual proofs of DK's algorithm, we found several problems.
The proof of Lemma 19 (Extension Equality Preservation) applies a wrong
induction hypothesis for the ${\longrightarrow}\mathtt{Uvar}$ case.
Fortunately the lemma could be proven without the well-formedness conditions
$\Gm\vdash A$ and $\Gm\vdash B$.
Similar problem happens for Lemma 14 (Subsumption), where many applications of
induction hypotheses are not correct.
For manual proofs, induction hypotheses are not automatically generated by programs,
thus are easily misused.

@@
